Incorporating Periodic Variability in Hidden Markov Models for Animal Movement

### 2016-08-10 bmc_template/figures.Rnw
BMB done for now

ICdat <- mutate(ICdat,
                model=factor(model,
                         levels=c("HMM","HMMsin"),
                         labels=c("Homogeneous","Heterogeneous"))) %>% filter(IC=="BIC")
simplot <- ggplot(ICdat,aes(x=nstates,y=n,color=model))+
                        geom_point(size=3) +
                        geom_line() +
                        scale_colour_brewer(palette="Set1")+
                        labs(y="Counts",x="BIC-optimal States")
print(simplot)

@@

<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles.",fig.width=6,fig.height=4>>=
allmodsumdf2 <- readRDS("../summary_stats/cat.1.df.RDS")$multimoddf
df2 <- transform(allmodsumdf2, #change allmodsumdf to allmodsumdf2
            distrib=ifelse(grepl("W",model),"Weibull","log-Normal"),
            turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
            homog=ifelse(grepl("TH",type),"Time-inhomogeneous HMM","Homogeneous HMM"))
(ggplot(df2, aes(x=nstates,y=deltaBIC,colour=distrib,lty=turn_angle))
    + geom_line()
    + facet_wrap(~ homog)
    + scale_colour_brewer(name="Step-length\ndistribution",palette="Set1")
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    + scale_linetype(name="von Mises\nturning angle?")
    + labs(y=expression(Delta*"BIC"),
         x="Number of States")
    + geom_hline(yintercept=0,colour="gray")
    + zmargin)
@

@@

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Adjusted BIC by number of free parameters for HMM model types. The left panel shows FMM, FMM with a sin prior and HMM. The right panel shows HMMs with different temporal transitions.",fig.width=8>>=
#L <- load("cat1sumdat.RData")

@@

mnames <- c("fmm5","fmmsin4","hmm6","hmmblock4","hmmhourly3",
            "hmmquad5","hmmsin5","obs","viterbi")
mnames2 <- c("fmm-homogeneous  (n=5)"
             , "fmm-sin (n=4)"
             , "homogeneous (n=6)"
             , "block (n=4)"
             , "hourly (n=3)"
             , "quadratic (n=5)"
             , "sin (n=5)"
             , "observed"
             , "viterbi homogenous (n=3)")

@@

all_cat_pars2 <- ldply(lme4:::namedList(1,2,14,15),get_cat_pars2)

g0 <- ggplot(all_cat_pars2,aes(log_mean,log_sd,color=Model))+
     geom_point()+geom_line() + facet_wrap(~catID,labeller=label_both)+zmargin
 (g0c <- g0 +
     geom_encircle(data=subset(all_cat_pars2,
                               (catID=="15" & log_mean<1.5) |
                               (catID!="15" & log_mean<1.8)),
                   colour="black",linetype=2,expand=0.05))

>>>>

ICdat <- mutate(ICdat,
                model=factor(model,
                         levels=c("HMM","HMMsin"),
                         labels=c("Homogeneous","Heterogeneous"))) %>% filter(IC=="BIC")
simplot <- ggplot(ICdat,aes(x=nstates,y=n,color=model,shape=model))+
    geom_line() +
    geom_point(size=3,bg="white") +
    scale_shape_manual(values=c(21,16))+
    scale_colour_brewer(palette="Set1")+
    labs(y="Counts",x="BIC-optimal number of states")
print(simplot)

@@

<<BICred_plot,warning=FALSE,echo=FALSE,message=FALSE,fig.cap="Adjusted Bayesian information criterion values for 3-7 state HMMs with different step-length distributions, with and without temporal transitions and turning angles.",fig.width=6,fig.height=4>>=
allmodsumdf2 <- readRDS("../summary_stats/cat.1.df.RDS")$multimoddf
df2 <- transform(allmodsumdf2, #change allmodsumdf to allmodsumdf2
            distrib=ifelse(grepl("W",model),"Weibull","log-Normal"),
            turn_angle=ifelse(grepl("VM",as.character(model)),"yes","no"),
            homog=ifelse(grepl("TH",type),"Time-heterogeneous","Homogeneous"))
slname <- "Step-length\ndistribution"
(ggplot(df2, aes(x=nstates,y=deltaBIC,size=distrib,
                 colour=distrib,lty=turn_angle))
    + geom_line()
    + facet_wrap(~ homog)
    + scale_colour_brewer(name=slname,palette="Set1")
    + scale_size_manual(name=slname,values=c(1.75,0.75))
    ## scale_size_continuous( name="# parameters") +
    #    geom_point(aes(size=parameters))+
    + scale_linetype(name="von Mises\nturning angle?")
    + labs(y=expression(Delta*"BIC"),
         x="Number of States")
    + geom_hline(yintercept=0,colour="gray")
    + zmargin)
@

@@

<<adj_BIC_comparisons, warning=FALSE, echo=FALSE, message=FALSE,fig.cap="Adjusted BIC by number of free parameters for HMM model types. The left panel shows FMM, temporally heterogeneous FMM with a sinusoidal prior and HMM. The right panel shows HMMs with different temporal transitions.",fig.width=8>>=
#L <- load("cat1sumdat.RData")

@@

mnames <- c("fmm5","fmmsin4","hmm6","hmmblock4","hmmhourly3",
            "hmmquad5","hmmsin5","obs","viterbi")
mnames2 <- c("FMM-homogeneous  (n=5)"
             , "FMM-sin (n=4)"
             , "HMM-homogeneous (n=6)"
             , "HMM-block (n=4)"
             , "HMM-hourly (n=3)"
             , "HMM-quadratic (n=5)"
             , "HMM-sin (n=5)"
             , "observed"
             , "viterbi homogenous (n=3)")

@@

all_cat_pars2 <- ldply(lme4:::namedList(1,2,14,15),get_cat_pars2)
levels(all_cat_pars2$Model) <- c("homogeneous","heterogeneous (sin)")
g0 <- ggplot(all_cat_pars2,
             aes(10^log_mean,log_sd*10^log_mean,color=Model,shape=Model))+
    scale_x_log10()+scale_y_log10()+
    labs(x="mean step length (m)",y="sd of step length (m)")+
    scale_shape_manual(values=c(21,16)) +
    geom_line() +
    geom_point(bg="white")+
    facet_wrap(~catID,labeller=label_both)+zmargin
 (g0c <- g0 +
     geom_encircle(data=subset(all_cat_pars2,
                               (catID=="15" & log_mean<1.5) |
                               (catID!="15" & log_mean<1.8)),
                   colour="black",linetype=2,expand=0.05,
                   group=1))
@

### 2016-08-10 bmc_template/manuscript.tex
BMB done for now

\bmb{doesn't this para. belong in methods??}
As a complement, we also fitted FMM and FMM with sinusoidal variation in state occupancy probabilities to compare the temporal effects in goodness of fit (dashed lines).
% \todo{I took out ``hourly HMM'' here - more coherent to describe it above} 
As a reminder, FMMs assume that the latent state in each time step is 
\emph{independent} of the latent state at the previous time step; 
time-varying FMMs can accurately describe movement when behaviour can
change on a short time scale, but the average propensity for different
behaviours changes over time.

@@

Models with temporal heterogeneity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogeneous HMMs are better than FMMs with sinusoidal temporal heterogeneity (Figure 3). Turning to the temporally heterogeneous HMMs (Figure 3, right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states. The multiple-block model approximately matches the homogeneous HMM, although it gives the BIC-optimal number of states as 4, in contrast to 6 for the homogeneous HMM.  Finally, the quadratic and sinusoidal models are considerably better than any other models tested; they both give the BIC-optimal number of states as 5, and they have similar goodness of fit.
However, this similarity is overstated due to the very large variation in BIC (over thousands of units) across the full range of models; there is a difference of approximately 80 BIC units, which would
normally be interpreted as an enormous difference in goodness of fit,
between the sinusoidal and quadratic models (both of which have 90 
parameters).

@@

Like the diurnal pattern (Figure 4),
the strong autocorrelation of the observed step lengths at a 24-hour lag (Figure 5) shows the need to incorporate temporal heterogeneity in the model --- we could have reached this conclusion even without developing any of the temporal-heterogeneity machinery.
Because there are a huge number of potential complexities that can be added to movement models (e.g. spatial/temporal/among-individual heterogeneity; effects of conspecific attraction or avoidance; memory or cognitive effects), each with associated costs in researcher and computational effort, such diagnostic plots are invaluable. \bmb{move prev sentence to Discussion??} In contrast to the hourly averages, the autocorrelation (ACF) captures both short- and long-term temporal effects. HMM without temporal heterogeneity captures the short-term autocorrelation, but misses the long-term autocorrelation beyond a 7-hour lag.
Temporally homogeneous FMM, by definition, produces neither short- nor long-term autocorrelation. FMM without temporal heterogeneity, although it captures the diurnal pattern well, underpredicts the degree of short-term autocorrelation.

The hardest problem with multiple latent states is interpreting them
biologically. We have no way of knowing what panthers are actually
thinking (it is certainly more complex than being in one of a small
number of discrete latent states); 
we don't know the ``true'' number of latent states,
nor are we able to observe them directly, although
incorporating additional direct observations of 
behaviour (if available) can at least partially
address this problem \cite{fryxell_multiple_2008}.
Three distinct
movement states seem biologically interpretable for Florida panthers
according to van de Kerk et al. \cite{kerk2015hidden}: Short step length suggests resting
states, intermediate step length a foraging state, and long
step length a traveling state.

The estimated parameter values
for several cats
(mean and standard deviation of the step length in each state) between
the time-homogeneous and time-heterogeneous models are similar across all cats (Figure 7). 
In general, the states with longer mean step lengths are relatively
similar between model classes. For cats 14 and 15, the states with
the longest or next-longest mean step lengths have similar means
and standard deviations; for cats 1 and 2, three long-step states
in the homogeneous HMM appear to divide two long-step states in 
the heterogeneous HMM. For short-step states, the heterogeneous HMM
tends to identify a high-variance state, while the homogeneous HMM
picks up states with very short step lengths (questionable in any
case because we have not taken any special efforts to account for
GPS error).

@@

\section*{Conclusion}

@@

\section{Data accessibility}
Hourly step lengths and turning angles of male and female Florida panthers available at: \url{http://ufdc.ufl.edu//IR00004241/00001}.

@@

\section*{Tables}
\begin{table}[h!]
\caption{Cat ID and number of observations; ID numbers are given matching
those shown by van de Kerk et al. 2014 and those 
in the data located at the UF Institutional repository (IR@UF).}
      \begin{tabular}{cccc}
        \hline
           & van de Kerk 2015  & IR@UF   & Number of Observations\\ \hline
         & 130 & 1 & 10286\\
         & 131 & 2  & 9458\\
         & 48 & 14  & 14645\\
         & 94  & 15   & 10250\\ \hline
      \end{tabular}
\end{table}

@@

  \begin{figure}[h!]
   \includegraphics[width=5in]{figure/adj_BIC_comparisons-1}
  \caption{\csentence{Adjusted BIC Transition Comparison} Adjusted BIC by number of free parameters for HMM model types. The left panel shows FMM, FMM with a sin prior and HMM. The right panel shows HMMs with different temporal transitions.}
      \end{figure}

@@

\begin{figure}[h!]
   \includegraphics[width=5in]{figure/r_msdlist-1}
  \caption{\csentence{State Identification Plot} Log mean and standard deviation of BIC-optimal HMM and HMM with sinusoidal transition.}
      \end{figure}

>>>>

\bmb{doesn't this para. belong in methods??}
As a complement, we also fitted FMM and FMM with priors on state occupancy that varied sinusoidally over time to compare the temporal effects in goodness of fit (dashed lines).
% \todo{I took out ``hourly HMM'' here - more coherent to describe it above} 
As a reminder, FMMs assume that the latent state in each time step is 
\emph{independent} of the latent state at the previous time step; 
time-varying FMMs can accurately describe movement when behaviour can
change on a short time scale, but the average propensity for different
behaviours changes over time.

@@

Models with temporal heterogeneity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogeneous HMMs are better than FMMs with sinusoidal temporal heterogeneity (Figure 3). Turning to the temporally heterogeneous HMMs (Figure 3, right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states. The multiple-block model gives approximately the same BIC as the homogeneous HMM, although it gives the BIC-optimal number of states as 4, in contrast to 6 for the homogeneous HMM.  Finally, the quadratic and sinusoidal models are the best models tested by far; they both give the BIC-optimal number of states as 5, and they have similar goodness of fit.
However, this similarity is overstated due to the very large variation in BIC (over thousands of units) across the full range of models; there is a difference of approximately 80 BIC units, which would
normally be interpreted as an enormous difference in goodness of fit,
between the sinusoidal and quadratic models (both of which have 90 
parameters).

@@

Like the diurnal pattern (Figure 4),
the strong autocorrelation of the observed step lengths at a 24-hour lag (Figure 5) shows the need to incorporate temporal heterogeneity in the model --- we could have reached this conclusion even without developing any of the temporal-heterogeneity machinery.
Because there are a huge number of potential complexities that can be added to movement models (e.g. spatial/temporal/among-individual heterogeneity; effects of conspecific attraction or avoidance; memory or cognitive effects), each with associated costs in researcher and computational effort, such diagnostic plots are invaluable. \bmb{move prev sentence to Discussion??} In contrast to the hourly averages, the autocorrelation (ACF) captures both short- and long-term temporal effects. HMM without temporal heterogeneity captures the short-term autocorrelation, but misses the long-term autocorrelation beyond a 7-hour lag.
Temporally homogeneous FMMs, by definition, produce no autocorrelation (neither short- nor long-term autocorrelation). FMMs without temporal heterogeneity, although they capture the diurnal pattern well, underpredict the degree of short-term autocorrelation.

\bmb{next para might belong in Discussion?}
The hardest problem with multiple latent states is interpreting them
biologically. We have no way of knowing what panthers are actually
thinking (it is certainly more complex than being in one of a small
number of discrete latent states); 
we don't know the ``true'' number of latent states,
nor are we able to observe them directly, although
incorporating additional direct observations of 
behaviour (if available) can at least partially
address this problem \cite{fryxell_multiple_2008}.
Three distinct
movement states seem biologically interpretable for Florida panthers
according to van de Kerk et al. \cite{kerk2015hidden}: Short step length suggests resting
states, intermediate step length a foraging state, and long
step length a traveling state.

The estimated emission parameter values
(mean and standard deviation of the step length in each state)
are similar, for both homogeneous and heterogeneous
models, across all cats (Figure 7 shows a subset of cats). 
In general, the states with longer mean step lengths are
similar between homogeneous and heterogeneous models. For cats 14 and 15, the states with
the longest or next-longest mean step lengths have similar means
and standard deviations; for cats 1 and 2, three long-step states
in the homogeneous HMM appear to divide two long-step states in 
the heterogeneous HMM. For short-step states, the heterogeneous HMM
tends to identify a high-variance state, while the homogeneous HMM
picks up states with very short step lengths (questionable in any
case because we have not taken any special efforts to account for
GPS error).

@@

\bmb{where did the BIC stuff (Celeux, finding the ``knee point''
à la Dean et al., etc.) go? I think we really have to mention this
(maybe not until discussion)}
\section*{Conclusion}

@@

\section{Data accessibility}
Hourly step lengths and turning angles of male and female Florida panthers available at \url{http://ufdc.ufl.edu//IR00004241/00001}.

@@

\section*{Tables}
\begin{table}[h!]
\caption{Cat ID and number of observations; ID numbers are given matching
those shown by van de Kerk et al. 2015 and those 
in the data located at the UF Institutional repository (IR@UF).}
\begin{tabular}{cccc}
  \hline
  & van de Kerk 2015  & IR@UF   & Number of Observations\\ \hline
  & 130 & 1 & 10286\\
  & 131 & 2  & 9458\\
  & 48 & 14  & 14645\\
  & 94  & 15   & 10250\\ \hline
\end{tabular}
\end{table}

@@

  \begin{figure}[h!]
   \includegraphics[width=5in]{figure/adj_BIC_comparisons-1}
  \caption{\csentence{Adjusted BICs Across All Models} Adjusted BIC by number of free parameters for HMM model types. The left panel shows homogeneous FMM, heterogeneous FMM (with a sinusoidal prior) and homogeneous HMM. The right panel shows HMMs with different temporal transitions.}
      \end{figure}

@@

\begin{figure}[h!]
   \includegraphics[width=5in]{figure/r_msdlist-1}
  \caption{\csentence{State Identification Plot} Mean and standard deviation of  step length by state for BIC-optimal HMMs (homogeneous and heterogeneous with sinusoidal transition). \bmb{do I need to explain transformation of LN SD ?}}
      \end{figure}
